Confluence in Lens Synthesis

### 2020-04-09 papers/BijectiveLenses/IWC/main.tex
minor edits

The judgement $\Lens : \Regex \Leftrightarrow \RegexAlt$ means that $\Lens$
bidirectionally converts between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. For synthesis, given $\Regex$ and $\RegexAlt$, we must
find a lens $\Lens$ such that $\Lens : \Regex \Leftrightarrow \RegexAlt$. There
are a number of typing rules for lenses, but the rules for type-equivalence and
composition (shown below) do not permit an efficient search algorithm. Note that
in type equivalence we use a finer notion of equivalence than language
equivalence, we instead use star-semiring equivalence. We were unable to make
our proof strategy work for language equivalence, so instead guarantee
completeness only on the subset of lenses well-typed with star-semiring
equivalence.
\begin{mathpar}
  \inferrule[\RewriteRegexLensRule{}]
  {
    \Lens \OfType \Regex \Leftrightarrow \RegexAlt\\
    \Regex \SSREquiv \Regex'\\
    \RegexAlt \SSREquiv \RegexAlt'
  }
  {
    \Lens \OfType \Regex' \Leftrightarrow \RegexAlt'
  }

@@

Our solution is to synthesize in an alternative language of disjunctive normal
form lenses (DNF lenses), typed by disjunctive number form regular expressions
(DNF REs). Both DNF lenses and DNF REs can be thought of as lenses and regular
expressions, put into a pseudonormal form.

@@

Two typing judgements are involved in typing a DNF lens, $\DNFLens : \DNFRegex
\Leftrightarrow \DNFRegexAlt$ and $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$. In both instances, the judgement means $\DNFLens$
is a DNF lens that converts between $\LanguageOf{\DNFRegex}$ and
$\LanguageOf{\DNFRegexAlt}$. If $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then the typing derivation of $\DNFLens$ never
rewrites the DNF REs. The only typing rule for $\DNFLens : \DNFRegex
\Leftrightarrow \DNFRegexAlt$ is:
\begin{mathpar}
  \inferrule[\RewriteDNFRegexLensRule{}]
  {
    \DNFRegex' \StarOf{\Rewrite} \DNFRegex\\
    \DNFRegexAlt' \StarOf{\Rewrite} \DNFRegexAlt \\
    \DNFLens \OfRewritelessType \MapsBetweenTypeOf{\DNFRegex}{\DNFRegexAlt}
  }
  {
    \DNFLens \OfType \MapsBetweenTypeOf{\DNFRegex'}{\DNFRegexAlt'}
  }
\end{mathpar}
To type a DNF lens, one first applies a number of rewrites to the DNF REs, then
builds up a DNF lens using rewriteless typing rules. Because of this, equivalent
types only need to be generated once, and composition never need be applied.
This permits an efficient algorithm that first searchest through type
equivalences with rewrites, and second builds up a rewriteless DNF lens purely
through the syntax of the DNF RE. This then permits the full synthesis
algorithm, shown in Figure~\ref{fig:schematic-diagram-synthesis}. The input
regular expressions are converted into DNF form with $\ToDNFRegex$. Next, a
well-typed DNF lens is synthesized, and then converted back into a standard lens
language with $\ToLens$.

Proving this search procedure complete reduces to proving DNF lenses complete
with respect to our standard lens language. In particular, we wish the prove the
following theorem:
\begin{theorem}
  If $\Lens : \Regex \Leftrightarrow \RegexAlt$, then there exists a DNF lens,
  $\DNFLens$, such that $\DNFLens \OfType \ToDNFRegexOf{\Regex} \Leftrightarrow
  \ToDNFRegexOf{\RegexAlt}$, such that the semantics of $\Lens$ and $\DNFLens$
  are equivalent.
\end{theorem}

We prove this by induction, and particular difficulty lies in the lens
equivalence rule. We begin this case below:

@@

The standard way of proving $(S,\StarOf{\rightarrow})$ to be $=$-confluent is to
first prove $(S,\rightarrow)$ to be $=$-confluent, then to perform two
inductions, first on the left arrows then on the right arrows. This approach
does not work for arbitrary $R$-confluence. Consider the inductive case, where
$R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $s_2 \rightarrow s_3$ and
$t_1 \StarOf{\rightarrow} t_2$, shown in Figure~\ref{fig:proof-attempt}. By
applying the inducion hypothesis, we know there exists $s_4, t_4$ such that $s_2
\rightarrow s_4$ and $t_2 \StarOf{\rightarrow} t_4$, and $R(s_4,t_4)$. We know
that $s_2 \rightarrow s_3$ and $s_2 \rightarrow s_4$, but we only know that
$(S,\rightarrow)$ is $R$-confluent, and so we are now stuck.

>>>>

The judgement $\Lens : \Regex \Leftrightarrow \RegexAlt$ means that $\Lens$
bidirectionally converts between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. For synthesis, given $\Regex$ and $\RegexAlt$, we must
find a lens $\Lens$ such that $\Lens : \Regex \Leftrightarrow \RegexAlt$. There
are a number of typing rules for lenses, but the rules for type-equivalence and
composition (shown below) do not permit an efficient search algorithm. Note that
in type equivalence we use a finer notion of equivalence than language
equivalence, we instead use star-semiring equivalence, denoted $\SSREquiv$. We were unable to make
our proof strategy work for language equivalence, so instead guarantee
completeness only on the subset of lenses well-typed with star-semiring
equivalence.
\begin{mathpar}
  \inferrule[\RewriteRegexLensRule{}]
  {
    \Lens \OfType \Regex \Leftrightarrow \RegexAlt\\
    \Regex \SSREquiv \Regex'\\
    \RegexAlt \SSREquiv \RegexAlt'
  }
  {
    \Lens \OfType \Regex' \Leftrightarrow \RegexAlt'
  }

@@

Our solution is to synthesize in an alternative language of disjunctive normal
form lenses (DNF lenses), typed by disjunctive normal form regular expressions
(DNF REs). Both DNF lenses and DNF REs can be thought of as lenses and regular
expressions, put into a pseudonormal form.

@@

Two typing judgements are involved in typing a DNF lens, $\DNFLens : \DNFRegex
\Leftrightarrow \DNFRegexAlt$ and $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$. In both instances, the judgement means $\DNFLens$
is a DNF lens that converts between $\LanguageOf{\DNFRegex}$ and
$\LanguageOf{\DNFRegexAlt}$. If $\DNFLens \OfRewritelessType \DNFRegex
\Leftrightarrow \DNFRegexAlt$, then the typing derivation of $\DNFLens$ never
rewrites the DNF REs. The only typing rule for $\DNFLens : \DNFRegex
\Leftrightarrow \DNFRegexAlt$ is:
\begin{mathpar}
  \inferrule[\RewriteDNFRegexLensRule{}]
  {
    \DNFRegex' \StarOf{\Rewrite} \DNFRegex\\
    \DNFRegexAlt' \StarOf{\Rewrite} \DNFRegexAlt \\
    \DNFLens \OfRewritelessType \MapsBetweenTypeOf{\DNFRegex}{\DNFRegexAlt}
  }
  {
    \DNFLens \OfType \MapsBetweenTypeOf{\DNFRegex'}{\DNFRegexAlt'}
  }
\end{mathpar}
To type a DNF lens, one first applies a number of rewrites to the DNF REs, then
builds up a DNF lens using rewriteless typing rules. Because of this
phase distinction, equivalent
types only need to be generated once, and composition never need be applied.
This structure corresponds to an efficient algorithm that first searches through type
equivalences with rewrites and then builds up a rewriteless DNF
lens as directed by the syntax of the DNF RE. 
Figure~\ref{fig:schematic-diagram-synthesis} shows the full synthesis
algorithm. The input
regular expressions are converted into DNF form using the
$\ToDNFRegex$ function. Next, a
well-typed DNF lens is synthesized and then converted back into a standard lens
language with the $\ToLens$ function.

Proving this search procedure complete reduces to proving DNF lenses complete
with respect to our standard lens language. In particular, we wish the prove the
following theorem:
\begin{theorem}
  If $\Lens : \Regex \Leftrightarrow \RegexAlt$, then there exists a DNF lens,
  $\DNFLens$, such that $\DNFLens \OfType \ToDNFRegexOf{\Regex} \Leftrightarrow
  \ToDNFRegexOf{\RegexAlt}$ and the semantics of $\Lens$ and $\DNFLens$
  are equivalent.
\end{theorem}

We prove this propety by induction. The particular difficulty lies in the lens
equivalence rule. We begin this case below:

@@

The standard way of proving $(S,\StarOf{\rightarrow})$ to be $=$-confluent is to
first prove $(S,\rightarrow)$ to be $=$-confluent, then to perform two
inductions, first on the left arrows then on the right arrows. This approach
does not work for arbitrary $R$-confluence. Consider the inductive case, where
$R(s_1,t_1)$ and $s_1 \StarOf{\rightarrow} s_2$ and $s_2 \rightarrow s_3$ and
$t_1 \StarOf{\rightarrow} t_2$, shown in Figure~\ref{fig:proof-attempt}. By
applying the induction hypothesis, we know there exists $s_4, t_4$ such that $s_2
\rightarrow s_4$ and $t_2 \StarOf{\rightarrow} t_4$, and $R(s_4,t_4)$. We know
that $s_2 \rightarrow s_3$ and $s_2 \rightarrow s_4$, but we only know that
$(S,\rightarrow)$ is $R$-confluent, and so we are now stuck.
